---
title: "R basics"
subtitle: ""  
author: ""
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

```{r xaringan-themer, include=FALSE}
library(xaringanthemer)
solarized_dark(
  code_font_family = "Fira Code",
  code_font_url    = "https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css"
)
```

background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)

???

Image credit: [Wikimedia Commons](https://commons.wikimedia.org/wiki/File:Sharingan_triple.svg)

---
name: intro_slide
class: center, middle
#Why use R

## - It is more than just a way to run statistics. you can make nice figures, run computer simulations, create interactive apps, and even make a website!

## - difference between a car and a bus

---
name: getting_help

#No one knows everything about R. i often Google the same issue many times since i forget.  A great place to look for help is **Stack Overflow**

---
name: setup

#Working with R

## The *console* is where commands are run. you can write something in the console to speak directly to R, or write in the *source* and then send the script to the console. You can recognize the console since every lines starts with a '>'

--

go to the console and type the following and then hit enter

```{r, echo = T, results='hide'}
7 + 6
```

---
you should see something like this:

> 7 + 6
[1] 13

note the [1], which tells you the line #. this becomes useful later on. also note it told you that 7 + 6 = 13. congrats! you just turned your expensive computer into a $5 calulator

---
name: how_R_thinks

#Short overview of how R works

##One way that i think about R that helps me to work it is to remember that everything in R is an "object."  we make objects (or  assign values to objects) and then do things to these objects.  

###for example check out the code below

```{r assign value}
a <- 4

```

what this says is that we are going to create the object 'a' and then assign the value '4' to it. if you run the code in R Studio you will see that the 'environment'  panel now lists 'a' and gives its value. 

???
what this says is that we are going to create the object 'a' and then assign the value '4' to it. if you run the code in R Studio you will see that the 'environment'  panel now lists 'a' and gives its value. 

---

#live examples
